---
title: "Cardiovascular Death Rate Change vs Air Quality Change"
author: "Jack Grafton"
date: "4/9/2018"
output: pdf_document
---

<!--***DAN: search for ***DAN in all docs to find my comments-->

<!--***DAN: This is a suitable idea for a final project, but so far it is not developed very far. You will have to take it further to make it a really good final project.-->

<!--DAN: Your grade on this unit testing assignment is 10/10, because the grade is based on effortful completion of all parts of the assignment. But that does not mean you did a perfect job - there are a number of areas for improvement, so please read the comments and let me know if anything is unclear.-->

#Introduction
This project analyzes the change in death rate from cardiovascular disease vs the change in Air Quality Index from 1980 to 2014, and how they may relate.

#Methods

<!--***DAN: No methods?-->

#Results
Preliminary result figures show two maps that represent change in aqi and change in cardiovascular death rate over 1980 to 2014.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#***DAN:This text should be outside a chunk, not in one.

# This proposal is for a project that analyzes the change in death from cardiovascular disease
# and compares it to the change in average aqi for each state, from the period 1980 to 2014.
```

<!--***DAN: need to cite where you got these data-->
```{r LoadData, echo=T}
aqi1980 <- read.csv("AQI_1980.csv")
aqi2014 <- read.csv("AQI_2014.csv")

mortChange <- read.csv("CardioMort1980_2014.csv")

```

```{r Function1_Source, echo=T}
source("Function1.R")
```

```{r CleanData, echo=T}
STaqi80 <- aggregate(MedianAQI ~ State, aqi1980, mean)
STaqi14 <- aggregate(MedianAQI ~ State, aqi2014, mean)


STaqi <- merge(STaqi80, STaqi14, by.x = "State", by.y = "State")

STaqiCR <- pctChange(STaqi, MedianAQI.x, MedianAQI.y)
# Using Function 1

STaqiChange <- aggregate(MedianAQI.z ~ State, STaqiCR, mean)

```

```{r Function3_Source, echo=T}
source("Function3.R")
```

```{r AnalyzeData, echo=T}
bigAQIchange <- Find_Significant(STaqiChange)
#Finds the states that had a significant change in aqi
bigMortChange <- Find_Significant(mortChange)


```


```{r PresentData, echo=T}
STaqiChange$region <- tolower(STaqiChange$State)
library(ggplot2)
library(maps)
states <- map_data("state")
map.df <- merge(states,STaqiChange, by="region", all.STaqiChange=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=MedianAQI.z))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map()

mortChange$region <- tolower(mortChange$State)
library(ggplot2)
library(maps)
states <- map_data("state")
map.df <- merge(states,mortChange, by="region", all.mortChange=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=MortalityChange))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+
  coord_map()

```

#Discussion
In a preliminary analysis, there seems to be some similarity between the change in cardiovascular death rate and the change in Air Quality Index. This similarity is especially evident in some of the southern states. While death dates have gone down, the rates that did not go down as much can be compared to aqi levels that have not improved much or gotten worse.

For the rest of the project, the plan is to more efficiently relate the two datasets in a meaningful way and display them in graphs as well as maps, and highlight key points where the death rate is not as improved in relation to AQI change.
